---
title: "Retrieving the Data"
output: rmarkdown::github_document
author: John Carlo Maula
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

This section focuses on retrieving the lyrics to Taylor Swift's albums using the genius package, which can be accessed [here](https://github.com/JosiahParry/genius). I will also be cleaning the data by removing duplicates, filling in missing values, and addressing contractions and special characters in the lyrics. Finally, I will store the dataset into a .csv file to be used in the analyses section.

**Note: Due to problems with the genius package, the resulting dataset from this section contains incomplete lyrics.**

### 1.1 Loading Packages

These are the packages I used for this section of the project.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Load required libraries
library(genius)
library(tidyverse)
```

### 1.2 Retrieving Lyrics

I used the genius library to pull the lyrics from Genius. I chose deluxe editions of albums where it is applicable to include any bonus tracks.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Create a tribble of inputs for each album
artist_albums <- tribble(
  ~artist, ~album,
  "Taylor Swift", "Taylor Swift",
  "Taylor Swift", "Fearless (Taylor's Version)",
  "Taylor Swift", "Speak Now (Deluxe)",
  "Taylor Swift", "Red (Deluxe Edition)",
  "Taylor Swift", "1989 (Deluxe)",
  "Taylor Swift", "reputation",
  "Taylor Swift", "Lover",
  "Taylor Swift", "folklore (deluxe version)",
  "Taylor Swift", "evermore (deluxe version)"
)

# Retrieve the lyrics for each album using the genius package
lyrics <- artist_albums %>% add_genius(artist, album, type = "album")
```


## 2. Data Cleaning

### 2.1 Examining the Data

After retrieving the lyrics from Genius, I viewed the data and its structure to determine any necessary cleaning. There are 6 columns and 3,552 rows. The column names are:

1. **artist** - artist of the song
2. **album** - album containing the song
3. **track_n** - track number of the song in the album
4. **line** - line number of the lyric in the song
5. **lyric** - lyric of the song
6. **track_title** - name of the song

**Note: Due to an issue with the genius package, the resulting lyrics are incomplete.**

```{r echo = TRUE, warning = FALSE, message=FALSE}
# View data
head(lyrics)

# Check structure of the data
str(lyrics)
```

### 2.2 Missing Values

There are two songs with missing values. It appears that the first lines of the song were mistaken as the second line, resulting in missing values for the first line. I removed these rows and corrected the line numbering. 

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Check missing values
lyrics[!complete.cases(lyrics), ]
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Check lyrics of tracks with the missing values
lyrics[lyrics$track_title == "I Knew You Were Trouble.", ]
lyrics[lyrics$track_title == "I Know Places", ]
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Remove rows with NA values
lyrics <- lyrics[complete.cases(lyrics), ]

# Fix line numbers for affected songs
lyrics[lyrics$track_title == "I Knew You Were Trouble.", ]$line <- lyrics[lyrics$track_title == "I Knew You Were Trouble.", ]$line - 1
lyrics[lyrics$track_title == "I Know Places", ]$line <- lyrics[lyrics$track_title == "I Know Places", ]$line - 1
```

### 2.3 Duplicated Lyrics

I looked at the list of every song in the dataset and verified the tracklisting. Then, I removed the extra versions of songs that are already in the album (e.g., *Teardrops on My Guitar (Pop Version)*).

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Check list of tracks included in the dataset
unique(as.factor(lyrics$track_title))
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Remove extra tracks
lyrics <- lyrics[(lyrics$track_title != "Teardrops on My Guitar (Pop Version)") &
                   (lyrics$track_title != "Forever & Always (Piano Version) [Taylor's Version]") &
                   (lyrics$track_title != "Love Story (Taylor's Version) [Elvira Remix]") &
                   (lyrics$track_title != "Treacherous (Original Demo Recording)") &
                   (lyrics$track_title != "Red (Original Demo Recording)") &
                   (lyrics$track_title != "State of Grace (Acoustic Version)") &
                   (lyrics$track_title != "I Know Places (Voice Memo)") &
                   (lyrics$track_title != "I Wish You Would (Voice Memo)") &
                   (lyrics$track_title != "Blank Space (Voice Memo)"), ]

```

### 2.4 Addressing Contractions and Special Characters

Using a [tutorial](https://www.datacamp.com/community/tutorials/R-nlp-machine-learning) from DataCamp, I defined a function that will fix contractions and remove special characters. Then, I applied these functions to the dataset and stored it into a csv file.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Check list of album titles
unique(lyrics$album) 

# Simplify album titles
lyrics$album <- gsub(" \\(Deluxe\\)", "", lyrics$album)
lyrics$album <- gsub(" \\(Deluxe Edition\\)", "", lyrics$album)
lyrics$album <- gsub(" \\(deluxe version\\)", "", lyrics$album)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Define function to fix contractions
fix.contractions <- function(doc) {
  # "won't" is a special case as it does not expand to "wo not"
  doc <- gsub("won't", "will not", doc)
  doc <- gsub("can't", "can not", doc)
  doc <- gsub("n't", " not", doc)
  doc <- gsub("'ll", " will", doc)
  doc <- gsub("'re", " are", doc)
  doc <- gsub("'ve", " have", doc)
  doc <- gsub("'m", " am", doc)
  doc <- gsub("'d", " would", doc)
  # 's could be 'is' or could be possessive: it has no expansion
  doc <- gsub("'s", "", doc)
  return(doc)
}

# Define function to remove special characters
remove.special.chars <- function(x) gsub("[^a-zA-Z0-9 ]", " ", x)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Apply functions to dataset
lyrics$lyric <- sapply(lyrics$lyric, tolower)
lyrics$lyric <- sapply(lyrics$lyric, fix.contractions)
lyrics$lyric <- sapply(lyrics$lyric, remove.special.chars)
```

Before storing the dataset into a csv file, I took one final look at the dataframe.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Preview the data and its structure
head(lyrics)

# Export dataset into a csv file
write.csv(lyrics, "data/swiftLyrics.csv")
```