---
title: "Sentiment Analysis"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

This section of the project focuses on analyzing the sentiments conveyed by Taylor Swift's lyrics. Many of the concepts here are inspired by the second chapter of the book *Text Mining with R: A Tidy Approach*, which can be accessed [here](https://www.tidytextmining.com/sentiment.html). I will be using three different lexicons, **Bing**, **NRC**, and **AFINN**, to analyze the sentiment of Taylor's music.

**Notes:** Updated on 02/09/2023 to include Taylor's new album, *Midnights*

### 1.1 Loading Packages

These are the packages I used for this section of the project.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Load required libraries
library(tidytext)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(gridExtra)
library(textdata)
```

### 1.2 Importing Data

I imported and previewed the data from *swiftLyrics_v2.csv*.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Importing the data
df <- read.csv("data/swiftLyrics_v2.csv")

# Preview the data
head(df)
```

### 1.3 Quick Data Preparation

I removed the extra column in the dataframe. I also performed some quick data preparation with the data including changing the lyrics into a character variable and parsing the lyrics into words and storing it into a dataframe.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Remove extra column "X"
df <- subset(df, select = -c(X))

# Change lyrics to character variable
df$Lyric <- as.character(df$Lyric)

# Turn strings of lyrics into words
words <- df %>% unnest_tokens(word,Lyric)

# Preview the words dataframe
head(words)
```

### 1.4 Defining Variables

I defined vectors of colors and color palettes that match the aesthetics of each album.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Define vector of album titles
album.list <- factor(c("Taylor Swift", "Fearless", "Speak Now", "Red", "1989", "reputation", "Lover", "folklore", "evermore", "Midnights"))

# Define vector of color hex codes associated with each album aesthetic
colors = c("#366E84", "#D9C78F", "#462245", "#A02B48", "#CDC6AC", "#333333", "#CC6B96", "#BABABA", "#67230E", "#154E52")
```

## 2. Sentimental Analysis

### 2.1 Positive vs. Negative Sentiment using the Bing Lexicon

The **bing lexicon** categorizes words into positive or negative categories. 

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Obtain the bing lexicon sentiments
bing <- get_sentiments("bing")

# Join the sentiments with the words dataframe
words_bing <- words %>% inner_join(bing)

# Preview joined dataframe
head(words_bing, 5)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Count the words for each sentiment
bing_word_counts <- words_bing %>% count(word, sentiment, sort = TRUE)

# Obtain top 10 positive and negative words
pos.counts <- bing_word_counts[bing_word_counts$sentiment == "positive", ][1:10, ]
neg.counts <- bing_word_counts[bing_word_counts$sentiment == "negative", ][1:10, ]

# Plot the word counts of the positive words
p1 <- ggplot(pos.counts, aes(reorder(word, n), n)) + 
  xlab("Word") + ylab("Count") + ggtitle("Positive Words") + 
  geom_col(fill = "lawngreen") + 
  geom_text(aes(label = n), hjust = 1.2, color = "black") +
  coord_flip() + 
  theme_classic()

# Plot the word counts of the negative words
p2 <- ggplot(neg.counts, aes(reorder(word, n), n)) + 
  ylim(0, 100) + xlab("Word") + ylab("Count") + ggtitle("Negative Words") + 
  geom_col(fill = "indianred1") + 
  geom_text(aes(label = n), hjust = 1.2, color = "black") +
  coord_flip() + 
  theme_classic()

grid.arrange(p1, p2, ncol=2)
```

According to the bing lexicon, words that contribute most to the positive sentiment of Taylor Swift's lyrics are **"like"**, **"love"**, **"right"**, and **"good"**. Words that contribute most to the negative sentiment are **"bad"**, **"break"**, **"shake"**, and **"lost"**. There are significantly more occurrences of positive words than negative words, with the top 3 positive words occurring 440, 251, and 124 times while the the top negative word occurs only 74 times. 

It's also interesting that the word **"shake"**, which has a negative association, appears the most frequently in her song *Shake It Off*, which is a positive, empowering song. 

### 2.2 Positive vs. Negative Sentiment using the AFINN Lexicon

The **AFINN lexicon** is similar to the bing lexicon except words are assigned a value between -5 and 5. Negative values indicate a negative sentiment and positive values indicate a positive sentiment.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Obtain the AFINN lexicon sentiments
afinn <- get_sentiments("afinn")

# Join the sentiments with the words dataframe
words_afinn <- words %>% inner_join(afinn)

# Preview joined dataframe
head(words_afinn)

# Calculate mean
mean(words_afinn$value)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Plot the distribution of AFINN values in Taylor's music
ggplot(words_afinn, aes(value)) +
  geom_bar(fill = "darkgreen", color = "black") +
  xlab("AFINN Value") +
  ylab("Word Count") +
  ggtitle("AFINN Values of Taylor Swift's Lyrics") +
  theme_classic()
```

Unsurprisingly, there are a lot more words with positive AFINN values in Taylor's lyrics than words with negative values with a mean AFINN value of 0.376.

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="50%"}
# Plot the distribution of AFINN values per album
for (i in 1:10) {
  p <- ggplot(words_afinn[words_afinn$Album == album.list[i], ], aes(value)) +
    geom_bar(fill = colors[i], color = "black") + 
    xlab("AFINN Value") + 
    ylab("Word Count") +
    ggtitle(album.list[i]) +
    theme_classic()
  
  print(p)
}
```

When it comes to individual albums, both *folklore* and *evermore* have more words with negative AFINN values with mean values of -0.052 and -0.043, respectively. Her most positive album is *Red* despite being an album about tumultuous emotions. It has a mean AFINN value of 0.632.

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="50%"}
# Calculate mean AFINN values per album
words_afinn %>% group_by(Album) %>% summarize(mean_afinn_value = mean(value))
```

## 2.3 Emotion Association with the NRC Lexicon

The **NRC lexicon** categorizes words in a binary fashion into categories of positive, negative, anger, anticipation, disgust, fear, joy, sadness, surprise, and trust. 

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Obtain the NRC lexicon sentiments
nrc <- get_sentiments("nrc") 

# Join the sentiments with the words dataframe
words_nrc <- words %>% inner_join(nrc)

# Preview the joined dataframe
head(words_nrc, 5)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Count the words for each sentiment
nrc_word_counts <- words_nrc %>% count(sentiment, sort = TRUE)

# Plot the NRC sentiment word counts
ggplot(nrc_word_counts, aes(reorder(sentiment, -n), n)) + 
  geom_col(aes(fill = sentiment)) + 
  xlab("Sentiment") + 
  ylab("Word Count") + 
  ggtitle("NRC Sentiment") + 
  scale_fill_brewer(palette="Spectral") + 
  geom_text(aes(label = n), vjust = 1.5, color = "black") +
  theme_classic() +
  theme(axis.text=element_text(size=12), axis.title=element_text(size=12), axis.text.x = element_text(angle = 30))
```

According to the NRC lexicon, the most common emotions conveyed by Taylor's lyrics are **positive**, **negative**, and **joy**. The least common emotion is **disgust**. 

```{r echo = TRUE, warning = FALSE, message=FALSE, fig.show="hold", out.width="50%"}
# Get nrc sentiment word counts by album
nrc_album_word_counts <- words_nrc %>% count(Album, sentiment, sort = TRUE)

# Plot AFINN Score per album
for (i in 1:10) {
  p <- ggplot(nrc_album_word_counts[nrc_album_word_counts$Album == album.list[i], ], aes(reorder(sentiment, -n), n)) + 
    geom_col(aes(fill = sentiment)) + 
    xlab("Sentiment") + 
    ylab("Word Count") + 
    ggtitle(album.list[i]) + 
    scale_fill_brewer(palette="Spectral") + 
    geom_text(aes(label = n), vjust = 1.5, color = "black") +
    theme_classic() +
    theme(axis.text=element_text(size=12), axis.title=element_text(size=12), axis.text.x = element_text(angle = 30))
  
  print(p)
}
```

When it comes to individual albums, the most common and least common emotions are generally consistent. There are, however, some slight changes that occur in the order of the emotions:

1. Words associated with **joy** are more common than **negative** words in the albums *Taylor Swift*, *1989*, and *Lover*.
2. Words associated with **anticipation** are common in *Speak Now*, *folklore*, and *evermore*.
3. Both *reputation* and *Lover* have more words associated with **disgust** than **surprise**.


## 3. Conclusion

This section focused on analyzing the sentiments conveyed by the lyrics of Taylor Swift's music using 3 different lexicons: bing, NRC, and AFINN. Both the bing and AFINN lexicon showed that her lyrics have a more positive sentiment than a negative sentiment. When it comes to individual albums, however, *folklore* and *evermore* convey a more negative sentiment.

The NRC lexicon associated words into 8 different emotions. Overall, the most common emotions associated with her lyrics are positive, negative, joy, and anticipation while the least common emotion is disgust. This result generally stays consistent throughout her albums with some minor differences.

However, these lexicons are not always correct in their interpretation. For example, the song *Shake It Off* is full of negative words such as "shake", "hate", "break" despite being an upbeat, positive song. Her album *Red* is considered the most positive album based on the AFINN lexicon despite being an album about tumultuous emotions. Despite this, these lexicons are mostly correct in categorizing the sentiment of Taylor's music.

Although it is known that Taylor Swift tends to write music about love, heartbreak, and relationships, this analysis shows that she tends to write in a more positive perspective about these topics. This contradicts the fairly common idea that she only writes songs to get back at her exes. I think it would be interesting to see how the Spotify feature "valence", which represents the musical positiveness conveyed by the track, is associated with these sentiments. It would also be interesting to see what emotions are associated with popular songs and see if this extends to music in general. For example, one can look at the sentiments of songs in the Billboard Hot 100 every week and determine if there is a pattern in the sentiments throughout the year. These sentiments have the potential to determine what emotions people are feeling based on the music they listen to.
