---
title: "Preparing the Dataset"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

Since retrieving the data using the genius library was returning incomplete lyrics, I decided to use a dataset, which I found [here](https://github.com/shaynak/taylor-swift-lyrics), that contains Taylor Swift lyrics. This section will focus on preparing that dataset for analysis, as well as addressing any other issues that occur such as missing lyrics. 

### 1.1 Loading Packages

These are the packages I used for this section of the project.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Load required libraries
library(tidyverse)
```

### 1.2 Importing Data

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Import lyrics data
lyrics <- read.csv("data/lyrics.csv")
```

## 2. Data Cleaning

### 2.1 Data Exploration

I previewed and looked at the structure of the data. There are 6 columns and 9,642 rows. The 6 columns are:

1. **Song** - the name of the song
2. **Album** - the album the song belongs to
3. **Lyric** - a line of the song
4. **Previous.Lyric** - the previous line of the song
5. **Next.Lyric** - the next line in the song
6. **Multiplicity** - the occurence of the lyric in the song

Since I only require the first 3 columns for this analysis, I dropped the last 3 columns.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Preview data
head(lyrics)

# Check data structure
str(lyrics)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Remove the last three columns
lyrics <- subset(lyrics, select = c("Song", "Album", "Lyric"))
```

For this project, I am only interested in songs that belong to her 9 studio albums: *Taylor Swift*, *Fearless*, *Speak Now*, *Red*, *1989*, *reputation*, *Lover*, *folklore*, and *evermore*.

Looking at the list of albums, there are 24 different album titles included in the dataset, such as the Hunger Games soundtrack. There are also deluxe versions of her studio albums, which contain bonus tracks. To simplify the album titles, I removed the "deluxe" identifiers in the album titles.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# View list of albums
levels(lyrics$Album)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Remove deluxe identifiers in the album titles
lyrics$Album <- gsub(" \\(Deluxe\\)", "", lyrics$Album)
lyrics$Album <- gsub(" \\(Deluxe Edition\\)", "", lyrics$Album)
lyrics$Album <- gsub(" \\(deluxe version\\)", "", lyrics$Album)
lyrics$Album <- gsub(" \\(Platinum Edition\\)", "", lyrics$Album)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Verify changes and change the album column to a factor variable
lyrics$Album <- factor(lyrics$Album)
levels(lyrics$Album)
```

### 2.2 Filtering Songs

I filtered the dataset to only contain lyrics from songs in her studio albums. After verifying the structure of the dataset, I found that there are 151 songs in the dataset.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Create a list of desired albums
albums <- c("Taylor Swift",
            "Fearless",
            "Speak Now",
            "Red",
            "1989",
            "reputation",
            "Lover",
            "folklore",
            "evermore")

# Filter songs in her nine studio albums
album.lyrics <- filter(lyrics, Album %in% albums) %>% droplevels()
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Verify 
str(album.lyrics)
```

There should be a total of 150 unique songs in the dataset. However, there are extra versions of songs that are included in the dataset (e.g., *Teardrops on My Guitar (Pop Version)*). Since these versions are usually indicated by an identifier in parenthesis, I decided to list all the songs that contain parenthesis in the title to identify the extra tracks to remove.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# List songs containing parenthesis in the title
par.songs <- album.lyrics[grep("\\(", album.lyrics$Song), ] %>% droplevels()
unique(par.songs$Song)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Remove extra tracks
album.lyrics <- album.lyrics[(album.lyrics$Song != "Teardrops on My Guitar (Pop Version)") &
                   (album.lyrics$Song != "Forever & Always (Piano Version)") &
                   (album.lyrics$Song != "Treacherous (Original Demo Recording)") &
                   (album.lyrics$Song != "Red (Original Demo Recording)") &
                   (album.lyrics$Song != "State of Grace (Acoustic Version)"), ] %>% droplevels()
```

After removing the extra versions of standard songs, there are a total of 146 songs remaining. Thus, there are a total of 4 missing songs in the dataset. Fortunately, they were already identified by the person who originally scraped the lyrics. 

The missing songs, which all belong to Taylor Swift's self-titled album, are:

1. Picture To Burn
2. Cold As You
3. Tied Together With A Smile
4. I'm Only Me When I'm With You

```{r echo = TRUE, warning = FALSE, message=FALSE}
# List songs in the self-titled album
unique(album.lyrics[album.lyrics$Album == "Taylor Swift", ]$Song)
```

### 2.4 Adding Missing Songs

Although I have incomplete lyrics of those songs from the data retrieval section, I decided to create a .txt file of the complete lyrics for each song and placed them into a designated folder. Then, I read these text files and stored them into a dataframe that matches the columns of the dataset. Finally, I appended the songs into the dataset.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Read extra tracks
picture.to.burn <- read.delim("data/missing_songs/picture_to_burn.txt", header = FALSE, col.names = "Lyric")
cold.as.you <- read.delim("data/missing_songs/cold_as_you.txt", header = FALSE, col.names = "Lyric")
tied.together <- read.delim("data/missing_songs/tied_together.txt", header = FALSE, col.names = "Lyric")
only.me <- read.delim("data/missing_songs/only_me.txt", header = FALSE, col.names = "Lyric")
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Add album and song titles
picture.to.burn$Song <- "Picture To Burn"
picture.to.burn$Album <- "Taylor Swift"

cold.as.you$Song <- "Cold As You"
cold.as.you$Album <- "Taylor Swift"

tied.together$Song <- "Tied Together With A Smile"
tied.together$Album <- "Taylor Swift"

only.me$Song <- "I'm Only Me When I'm With You"
only.me$Album <- "Taylor Swift"
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Add songs to the dataset
complete.album.lyrics <- rbind(album.lyrics, picture.to.burn, cold.as.you, tied.together, only.me)
```

### 2.5 Special Characters & Contractions

Using a [tutorial](https://www.datacamp.com/community/tutorials/R-nlp-machine-learning) from DataCamp, I defined a function that will fix contractions and remove special characters. Then, I applied these functions to the dataset and stored it into a csv file.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Define function to fix contractions
fix.contractions <- function(doc) {
  # "won't" is a special case as it does not expand to "wo not"
  doc <- gsub("won't", "will not", doc)
  doc <- gsub("can't", "can not", doc)
  doc <- gsub("n't", " not", doc)
  doc <- gsub("'ll", " will", doc)
  doc <- gsub("'re", " are", doc)
  doc <- gsub("'ve", " have", doc)
  doc <- gsub("'m", " am", doc)
  doc <- gsub("'d", " would", doc)
  # 's could be 'is' or could be possessive: it has no expansion
  doc <- gsub("'s", "", doc)
  return(doc)
}

# Define function to remove special characters
remove.special.chars <- function(x) gsub("[^a-zA-Z0-9 ]", " ", x)
```

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Apply functions to dataset
complete.album.lyrics$Lyric <- sapply(complete.album.lyrics$Lyric, tolower)
complete.album.lyrics$Lyric <- sapply(complete.album.lyrics$Lyric, fix.contractions)
complete.album.lyrics$Lyric <- sapply(complete.album.lyrics$Lyric, remove.special.chars)
```

Before storing the dataset into a csv file, I took one final look at the dataframe to make sure the results are as expected.

```{r echo = TRUE, warning = FALSE, message=FALSE}
# Preview the data 
head(complete.album.lyrics, 10)

# Export dataset into a csv file
write.csv(complete.album.lyrics, "data/swiftLyrics_v2.csv")
```